# Examples {#examples}

## Load the libraries and the npde package

```{r setup, echo=TRUE, message=FALSE}
# Libraries
library( gridExtra )
library( ggplot2 )
library( grid )
library( devtools )
library( mclust )
library( npde )
```
## Load the data and the simulated data

Github repository for the data
<https://github.com/ecomets/npde30/tree/main/keep/data>

```{r warfarin_data, echo=TRUE}
# Data 
data( warfarin )
data( simwarfarinCov )

fullDatDir = paste0(getwd(),"/data/")

simwarfarinBase = read.table( file.path( fullDatDir, "simwarfarinBase.tab" ), header = TRUE )
```
### Warfarin : description of the data

```{r warfarin_description, echo=TRUE }
# Warfarin
head( warfarin )

# Warfarin with base model
head( simwarfarinBase )

# Warfarin with covariate model
head( simwarfarinCov )
```
## Compute the normalised prediction distribution errors

```{r warfarin_autonpde, echo=TRUE, warning=FALSE}
wbase <- autonpde( namobs = warfarin, namsim = simwarfarinBase, 
                   iid = 1, ix = 2, iy = 4, icov = c( 3,6:8 ), namsav = "warfBase", 
                   units = list( x = "hr", y = "ug/L", covariates = c( "mg","kg","-","yr" ) ) )

wcov <- autonpde( namobs = warfarin, namsim = simwarfarinCov, 
                  iid = 1, ix = 2, iy = 4, icov = c( 3,6:8 ), namsav = "warfCov", 
                  units = list( x = "h", y = "mg/L", covariates = c( "mg","kg","-","yr" ) ) )
```

```{r warfarin_autonpde_output, echo=TRUE, warning=FALSE}
show( wbase )
print( wbase )
summary( wbase )
```

### Description of the output wcov

```{r warfarin_cov_autonpde_description, echo=TRUE}

# wcov
head( wcov )

# wcov slots
slotNames( wcov )

# wcov covariates
wcov@data@name.covariates

# wcov graphical options
#wcov@prefs

```

## Data 

### Plot the data for the base model

``` plot.type = "data" ``` : Plots the observed data in the dataset

```{r warfarin_wbase_plot_data, echo=TRUE}
# wbase model
plot( wbase, plot.type = "data" )
```

### Default plot for the base model

```{r warfarin_base_plot_default, echo=TRUE}
# base model
plot( wbase, which = "npde" )
```

``` plot.type = "hist" ``` Histogram  

``` plot.type = qqplot``` QQ-plot of the npde versus its theoretical distribution 

``` plot.type = x.scatter``` Scatterplot of the npde versus the predictor 

``` plot.type = pred.scatter``` Scatterplot of the npde versus the population predicted values 

```{r warfarin_base_plot_default_each, echo=TRUE}
hist = plot( wbase, plot.type = "hist", which = "npde" )

qqplot = plot( wbase, plot.type = "qqplot", which = "npde" )

x_scatter = plot( wbase, plot.type = "x.scatter", which = "npde" )

pred_scatter = plot( wbase, plot.type = "pred.scatter", which = "npde" )

grid.arrange( grobs = list( hist, qqplot ), nrow = 1, ncol = 2 )

grid.arrange( grobs = list( x_scatter, pred_scatter ), nrow = 1, ncol = 2 )

```

## Covariate model

### Default plots 

```{r warfarin_cov_plot_default, echo=TRUE}
# covariate model
plot( wcov , which = "npde")
```

### Scatterplots of npd

``` plot.type = "x.scatter" ``` : Scatterplot of the npd versus the predictor 

``` covsplit = TRUE ```  : split by categories

```{r warfarin_cov_scatter_split_covariates, echo=TRUE, message = FALSE}

# Splitting npde vs time plots by covariates
plot( wcov, plot.type = "x.scatter", covsplit = TRUE, which.cov = c( "wt" ) )
# plot( wcov, plot.type = "x.scatter", covsplit = TRUE, which.cov = c( "wt","sex" ) )
# plot( wcov, plot.type = "x.scatter", covsplit = TRUE, which.cov = c( "wt" ), which = "pd" )
# plot( wcov, plot.type = "x.scatter", covsplit = TRUE, which.cov = c( "wt" ), which = "npde" )

```

### Plot npd as boxplots for each covariate category

``` plot.type = "covariates" ``` npd represented as boxplots for each covariate category 

```{r warfarin_cov_scatter_split_covariates_box_plot, echo=TRUE, warning=FALSE}

wt.covariate.boxplot <- plot( wcov, plot.type = "covariates", which.cov = c( "wt" ) )
sex.covariate.boxplot <- plot( wcov, plot.type = "covariates", which.cov = c( "sex" ) )

# grid plot
grid.arrange( grobs = list( wt.covariate.boxplot[[1]], sex.covariate.boxplot[[1]] ), 
              nrow = 1, ncol = 2 )

```


### Diagnostic plots for covariates


``` plot.type = "cov.scatter" ``` : Scatterplot of the npd versus covariates 

``` plot.type = "covariates" ``` : Boxplots for each covariate category

``` plot.type = "ecdf" ``` : Empirical distribution function 

```{r warfarinCovPlotDiagnostic, echo=TRUE, warning=FALSE}

# cov.scatter
xwt.covscatt <- plot( wcov, plot.type = "cov.scatter", which.cov = "wt", bin.method = "optimal" )

# ecdf
xwt.ecdf <- plot( wcov, plot.type = "ecdf", covsplit = TRUE, which.cov = "wt" )

# x.scatter
xsex.covscatt <-  plot( wcov, plot.type = "x.scatter", covsplit = TRUE, which.cov = "sex" )

# ecdf
xsex.ecdf <- plot( wcov, plot.type = "ecdf", covsplit = TRUE, which.cov = "sex" )

# grid plot
grid.arrange( grobs = list( xwt.covscatt, xwt.ecdf, xsex.covscatt, xsex.ecdf ), 
              nrow = 2, ncol = 2 )

```

## Reference profile

### Reference plot using one subject

```{r warfarin_cov_reference_profil_one_subject, echo=TRUE, warning=FALSE}

plot.tnpde <- plot( wcov, plot.type = "x.scatter", 
                    ref.prof = list( id = 2 ), 
                    main = "tnpd with reference profile ID = 2" )

plot.vpc <- plot( wcov, 
                  plot.type = "vpc", 
                  main = "VPC" ) 

grid.arrange( grobs = list( plot.tnpde, plot.vpc ), nrow = 1, ncol = 2 )

```

### Reference plot using all subjects  

```{r warfarin_cov_reference_profil_all_subject, echo=TRUE, warning=FALSE}

plot.tnpde <- plot( wcov, plot.type = "x.scatter", 
                    ref.prof = "all", main = "tnpd", ylim = c ( 0,20 ) )

plot.vpc <- plot( wcov, plot.type = "vpc", main = "VPC", ylim = c( 0, 20 ) )

grid.arrange( grobs = list( plot.tnpde, plot.vpc ), nrow = 1, ncol = 2 )

```


  
## Computing npde in the presence of BQL data  
  
```{r virloadData, echo=TRUE}

data( virload )
data( virload20 )
data( virload50 )

virload <- read.table( file.path( fullDatDir, "virload.tab" ), header = TRUE )
simvirload <- read.table( file.path( fullDatDir, "simvirload.tab" ), header = TRUE )

```
#### Censoring methods

``` cens.method = "omit" ```

``` cens.method = "ipred" ```

``` cens.method = "ppred" ```

``` cens.method = "cdf" ```


```{r virloadData_autonpde, echo=TRUE}

x50 <- autonpde( virload50, simvirload, 
                 iid = 1, ix = 2, iy = 3, icens = 4,
                 boolsave = FALSE, units = list( x = "hr", y = "cp/mL, log 10 base" ) )

x50.ipred <- autonpde( virload50, simvirload,
                       iid = 1, ix = 2, iy = 3, icens = 4, iipred = 5, 
                       boolsave = FALSE, cens.method = "ipred", 
                       units = list( x = "hr", y = "cp/mL, log 10 base" ) )

x50.ppred <- autonpde( virload50, simvirload,
                       iid = 1, ix = 2, iy = 3, icens = 4,
                       boolsave = FALSE, cens.method = "ppred", 
                       units = list(x = "hr", y = "cp/mL, log 10 base" ) )

x50.omit <- autonpde( virload50, simvirload,
                      iid = 1, ix = 2, iy = 3, icens = 4,
                      boolsave = FALSE, cens.method = "omit", 
                      units = list( x = "hr", y = "cp/mL, log 10 base" ) )

```

### Data

``` impute.loq ``` : the imputed values are plotted for data under the LOQ

``` line.loq ``` : horizontal line should is plotted at Y=LOQ in data and VPC plots

``` plot.loq ``` : data under the LOQ are plotted

```{r }
x50@data@loq
```

```{r virloadDataPlotCensored, echo=TRUE}

x1 <- plot( x50, plot.type = "data", 
            xlab = "Time (hr)", ylab = "log(Viral load) (cp/mL)", 
            line.loq = TRUE, ylim = c(0,6.5), 
            main = "LOQ imputed using cdf" )

x2 <- plot( x50, plot.type = "data", 
            xlab = "Time (hr)", ylab = "log(Viral load) (cp/mL)", 
            plot.loq = FALSE, line.loq = TRUE, ylim = c(0,6.5), 
            main = "LOQ removed from plot" )

x3 <- plot( x50, plot.type = "data", 
            xlab = "Time (hr)", ylab = "log(Viral load) (cp/mL)", 
            impute.loq = FALSE, line.loq = TRUE, ylim = c(0,6.5),
            main = "LOQ as in dataset before imputation" )

x4 <- plot( x50.ipred, plot.type = "data", 
            xlab = "Time (hr)", ylab = "log(Viral load) (cp/mL)", 
            line.loq = TRUE, ylim = c(0,6.5), 
            main = "LOQ imputed to individual prediction" )

grid.arrange( grobs = list(x1,x2,x3,x4 ), nrow = 2, ncol = 2 )

```
  
### Comparing the censoring methods
  
#### Default graphs for the virload50 dataset, default censoring method "cdf"
```{r docVirloadCensoringMethods_cdf, echo=FALSE}
plot(x50)
```

#### Default graphs for the virload50 dataset, censoring method "omit"
```{r docVirloadCensoringMethods_omit, echo=FALSE}
plot(x50.omit)
```

#### Default graphs for the virload50 dataset, censoring method "ipred"
```{r docVirloadCensoringMethods_ipred, echo=FALSE}
# Default graphs for the virload50 dataset, censoring method "ipred".
plot(x50.ipred)
```

#### Default graphs for the virload50 dataset, censoring method "ppred"
```{r docVirloadCensoringMethods_ppred, echo=FALSE}
plot(x50.ppred)
```

### VPC plots

```{r docVirloadVPC, echo=TRUE, warning=FALSE}

vpc.cdf = plot( x50,plot.type = "vpc", line.loq=TRUE )

vpc.omit = plot( x50.omit,plot.type = "vpc" )

grid.arrange( grobs = list( vpc.omit, vpc.cdf ), nrow=1, ncol=2)

```

### Scatterplots

```{r docVirloadScatterPlot, echo=TRUE, warning=FALSE}

xscatter.omit <- plot( x50.omit, plot.type = "x.scatter" )

xscatter.cdf <- plot( x50, plot.type = "x.scatter" )

grid.arrange( grobs = list( xscatter.omit, xscatter.cdf ), nrow = 1, ncol = 2 )

plot(x50 , plot.type = "x.scatter", plot.box = TRUE )

```

### Plot for $\mathbb{P}(Y<LOQ)$

```{r docVirloadPInfLOQ, echo=TRUE, warning=FALSE}

# P(Y<LOQ)

plot( x50, plot.type = "loq", main = " ")

```

 
